\textbf{Effective application of machine learning to bacterial 16S rRNA gene sequencing data}

### 2019-09-11 submission/manuscript.tex
version sent to Jenna

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
diseases. However, the structure of the human microbiome is remarkably
variable between individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population individually. Recently,
machine learning models have grown in popularity among microbiome
researchers because of the large amount of data that can now be
generated and because the models are effective at accounting for the
interpersonal microbiome variation and the ecology of the disease.

ML models are useful for understanding the variation in the structure of
existing data and to apply that knowledge to make predictions about new
data. Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, and type 2 diabetes (1--16,
16--18). The task of diagnosing an individual with high confidence
relies on a model that is built with rigorous methods. However, there
are common methodological problems across many of these studies that
need to be addressed as the field progresses. These include a lack of
transparency in which methods are used and how these methods are
implemented; developing and evaluating models without a separate
held-out test data; large variation between the predictive performance
on different folds of cross-validation; and large variation between
cross-validation and testing performances. Nevertheless, the microbiome
field is making progress to avoid some of these pitfalls including
validating their models on independent datasets (7, 18, 19) and
introducing analysis frameworks to better use ML tools (20--23). More
work is needed to further improve reproducibility and minimize
over-optimism for model performance.

@@

\textbf{Model selection and pipeline construction} We established a ML
pipeline where we trained and validated each of the seven models using a
common approach that is based on standard methods within the ML
community (REFS by Jenna?){[}Figure 1{]}.

@@

Second, we trained seven different models using the training data
{[}Table 1{]}. We selected models with different classification
algorithms and regularization methods. Regularization is a technique
that discourages overfitting by penalizing the model for learning the
training data too well. For regularized logistic regression and SVM with
linear kernel, we used L2 regularization to keep all potentially
important features. For comparison, we also trained an L1-regularized
SVM model with linear kernel. L1-regularization on microbiome data led
to a sparser solution (i.e., force many coefficients to zero). To
explore the potential for non-linear relationships among features to
improve classification, we trained tree-based models including decision
tree, random forest, and XGBoost and we trained an SVM model with
non-linear kernel.

Third, fitting of these models require selecting appropriate
hyperparameters. Hyperparameters are the rules that are learned from the
training set in a classification algorithm. For example, in the linear
models the regularization term (C) is a hyperparameter that indicates
the penalty for overfitting. Similar to regularization term C, all
hyperparameters are tuned to find the best model. We selected
hyperparameters by performing 100 five-fold cross-validation (CV)
repeats on the training set {[}Figure 1{]}. The five-fold CV was also
stratified to maintain the overall case and control distribution. We
chose the best hyperparameter values for each model based on its CV
predictive performance using the area under the receiver operating
characteristic curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC
ranges from 1.0, where the model perfectly distinguishes between cases
and controls, to 0.50, where the model's predictions are no different
from random chance. To select the best performing hyperparameter, we
performed a full grid search for hyperparameter settings when training
our models. Default hyperparameter settings in previously developed ML
packages in R, Python, and Matlab programming languages are inadequate
for effective application of classification algorithms and need to be
optimized for each new ML task. In the example of L1-regularized SVM
with linear kernel {[}Figure S1{]}, the model showed large variability
between different regularization coefficients (C) and was susceptible to
performing poorly if the wrong regularization coefficient was assigned
to the model by default.

Finally, we trained the full training dataset with the selected
hyperparameter values and applied the model to the held-out data to
evaluate the testing predictive performance of each model. The
data-split, hyperparameter selection, training and testing steps were
repeated 100 times to get a reliable and robust reading of model
performance {[}Figure 1{]}.

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had significantly higher test AUROC values than the
other models for detecting SRNs (Wilcoxon rank sum test, p \textless{}
0.01). The median AUROC of the random forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another and had median AUROC values of
0.68 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056), respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.65 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Interestingly, these results demonstrate that the most
complex model (XGBoost) did not have the best performance and that the
most interpretable models (L2-regularized logistic regression and SVM
with linear kernel) performed nearly as well as random forest.

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs was large, then that
would indicate that the models were overfit. The difference in median
AUROCs was 0.021 in L1-regularized SVM with linear kernel, followed by
SVM with radial basis function kernel and decision tree with a
difference of 0.007 and 0.006, respectively {[}Figure 2{]}; however,
these differences are relatively small and would not indicate a problem
with overfitting.

To evaluate the risk for over-optimism of each model, we calculated the
range of AUROC values for each model using 100 data-splits. The range
among the testing AUROC values within each model varied by 0.23 on
average across the seven models. If we had only done a single split,
then there is a risk that we could gotten lucky or unlucky with the
performance of the model. For instance, the lowest AUROC value of the
random forest model was 0.593 whereas the highest was 0.81. These
results showed that depending on the data-split, the testing AUROC
values showed great variability {[}Figure 2{]}. Therefore, it is
important to employ the hierarchical data splits that were included in
our pipeline to minimize the risk of over-optimism.

@@

\textbf{Interpretation of each ML model.} Interpretability is the degree
to which humans can understand the reasons behind a model prediction
(31). Because we often use ML models not just to predict a health
outcome but also to learn the ecology behind a disease, model
interpretation becomes crucial for microbiome studies. ML models
decrease in interpretability as they increase in complexity. In this
study we used two methods to help interpret our models.

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU is associated with classifying a subject as
being healthy or having an SRN - negative sign indicated being healthy
and positive sign indicated having an SRN. It was encouraging that many
of the highest ranked OTUs were shared across these three models,
(e.g.~OTU 50, 426, 609, 822, 1239). The benefit of this approach was
that the results of the analysis were based on the trained model
parameters and provided information regarding the sign and magnitude of
the impact of each OTU. However, this approach was only possible with
linear models.

Second, to analyze non-linear models we interpreted the feature
importance using permutation importance. Whereas the absolute feature
weights were determined from the trained models, here we measured
importance using the held-out test data. Permutation importance analysis
is a posthoc explanation of the model where we randomly permuted
non-correlated features individually and groups of perfectly correlated
features across the two groups in the held-out test data. We then
calculated how much the predictive performance of the model (i.e testing
AUROC values) decreased when each OTU or group of OTUs was randomly
permuted. We ranked the OTUs based on how much the median testing AUROC
decreased when it was permuted; the OTU with the largest decrease ranked
highest {[}Figure 4{]}. Among the twenty OTUs with the largest impact,
there was only one OTU (OTU 822) that was shared among all of the
models; however, we found three OTUs (OTU 58, 110, 367) that were
important in each of the tree-based models. Similarly, the random forest
and XGBoost models, shared four of the most important OTUs (OTU 2, 12,
361, 477). Permutation analysis results also revealed that with the
exception of the decision tree model, removal of any individual OTU had
a minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.68.
Effectively, the complexity of the communities was more fully
represented in the better performing models (22, 32). At least in this
case, it was not possible to distinguish between health and disease
using a single OTU. While permutation analysis allowed us to gauge the
importance of an OTU, the analysis was post-hoc (i.e.~done using the
test data) and these results did not allow us to directly interrogate
the models to know whether an OTU is associated with classifying a
subject as being healthy or having an SRN.

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S4{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S4{]}, 17 out of the 20 top OTUs
(e.g.~OTU 609, 822, 1239) were deemed important by both interpretation
methods. Similarly, for the L2-regularized SVM and L2-regularized
logistic regression, 9 and 12 OTUs, respectively, were shared among the
two interpretation methods. These results indicate that both methods are
consistent in selecting the most important OTUs.

@@

There is a growing awareness that many human diseases and environmental
processes are not driven by a single organism but are the product of
multiple bacterial populations. Traditional statistical approaches are
useful for identifying those cases where a single organism is associated
with a process. In contrast, ML methods offer the ability to incorporate
the structure of the microbial communities as a whole to classify them
into different categories such as coming from a patient who is healthy
or has SRNs. If it is possible to classify communities reliably, then ML
methods also offer the ability to identify those microbial populations
within the communities that are responsible for the classification.
However, the application of ML in microbiome studies is still in its
infancy and the field still needs to develop a better understanding of
different ML methods, their strengths and weaknesses, and how to
implement them.

To address these needs, we developed a framework to train rigorous,
transparent, and reproducible models. We benchmarked seven ML models and
showed that we can create models that are inherently interpretable and
easily trained without losing predictive performance. In terms of
predictive performance, the random forest model had the best testing
AUROC values compared to the other six models. However, the second-best
model was L2-regularized logistic regression with a median AUROC
difference of only 0.015 compared to random forest. While random forest
took 83.2 hours to train, L2-regularized logistic regression trained in
12 minutes. In terms of interpretability, random forest was a complex ML
model and could only be explained using post-hoc methods such as
permutation importance. On the other hand, L2-regularized logistic
regression was easier to interpret by ranking the OTUs based on their
feature weights and reviewing the signs of these weights. Comparing many
different models showed us that the most complex model is not
necessarily the best model for our ML task.

As we set out to select the best model, we established a pipeline that
can be generalized to any modeling method that uses 16S rRNA sequence
counts to predict a binary health outcome. We performed a random
datasplit to create a training set (80\% of the data) and a held-out
test set (20\% of the data), which we used to evaluate predictive
performance. We repeated this datasplit 100 times to measure predictive
performance. During the training, we tuned the model hyperparameters
with a repeated five-fold cross-validation. Despite the high number of
features microbiome datasets typically have, the models we built with
this pipeline were generalizable as shown by the similar AUROC values
from the cross-validation and testing.

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures which create collinearity in the datasets we work with. This
can hinder our ability to reliably interpret models because the feature
weights of correlated OTUs are influenced by one another (33). For
example if one OTU is highly correlated with another OTU, only one of
these OTUs would have a large feature weight thus hiding the importance
of its correlated OTU. To capture all important features, once we
identify highly ranked OTUs, we should review their relationships with
other OTUs. These relationships will help us generate new hypotheses
about the ecology of the disease and test them with follow-up
experiments. When we used permutation importance, we took collinearity
into consideration by grouping correlated OTUs to determine their impact
as a group. We grouped OTUs that had a perfect correlation with each
other however, we can reduce the correlation threshold to further
investigate the relationships among features. It is important to know
the correlation strutures of the data to avoid missinterpreting the
models. This is likely to be a particular problem with shotgun
metagenomic datasets where collinearity will be more pronunced due to
many genes being correlated with one another because they come from the
same chromosome. To identify the true underlying microbial factors of a
disease, it is crucial to do correlation analyses and further
experimentation for biological validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to all
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. For
example, we could have trained models to differentiate between people
with normal colons and those with adenomas or carcinomas (k=3
categories). We did not perform this analysis because the clinically
relevant diagnosis grouping was between patients with normal colons and
those with SRNs. Furthermore, as number of categories to classify
increase, more samples are required for each category to train a model.
We also did not use regression-based analyses to predict a
non-categorical outcome. We have previously used such an approach to
train random forest models to predict fecal short-chain fatty acid
concentrations based on microbiome data (34). Our analysis was also
limited to shallow learning methods and did not explore deep learning
methods such as neural networks. Deep learning methods hold great
promise (11, 35, 36) but microbiome datasets often suffer from having
many features and small sample sizes, which makes the deep learning
models prone to overfitting. These methods are even more complex than
random forest and XGBoost and are considered uninterpretable. There is
great potential for applying ML approaches to microbiome data and we can
choose which ML approach is best for the study.

@@

This study highlighted the need to make educated choices at every step
of developing a ML model with microbiome data. We created an
aspirational rubric that researchers can use to identify potential
pitfalls when using ML in microbiome studies and ways to avoid them
{[}Table S1{]}. We have highlighted the tradeoffs between model
complexity and interpretability, the need for cross-validation to tune
hyperparameters, the utility of held-out test sets for evaluating
predictive performance, and the importance of considering correlation
structures in datasets for reliable interpretation. Furthermore, we
underscored the importance of proper experimental design and methods to
help us achieve the level of validity and accountability we want from
models built for patient health.

@@

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (Reference
from Mack?). Stool samples were provided by adults who were undergoing a
scheduled screening or surveillance colonoscopy. Participants were
recruited from four locations: Toronto (ON, Canada), Boston (MA, USA),
Houston (TX, USA), and Ann Arbor (MI, USA). Patients' colonic health was
visually assessed by colonoscopy with bowel preparation and tissue
histopathology of all resected lesions. We assigned patients into two
classes: those with normal colons and those with screen relevant
neoplasias (SRNs). The normal class included patients with normal colons
or non-advanced adenomas whereas the SRN class included patients with
advanced adenomas or carcinomas (reference for SRN?). Patients with an
adenoma greater than 1 cm, more than three adenomas of any size, or an
adenoma with villous histology were classified as having advanced
adenomas. The study had 172 patients with normal colonoscopies, 198 with
adenomas, and 120 with carcinomas. Of the 198 adenomas, 109 were
identified as advanced adenomas. The study collected samples from 261
patients classified as normal and 229 patients classified as having a
SRN.

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients was used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, 2015). The OTU abundance table was
generated by Sze et al (32), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (37--39);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1 and
lowest would be 0.

\textbf{Model training and evaluation.} Models were trained using the
caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to
calculate decision values instead of predicted probabilities for models
generated using L2-regularized SVM with linear kernel and L1-regularized
SVM with linear kernel. These changes were necessary to calculate AUROC
values for SVMs. The code for these changes on L2-regularized SVM with
linear kernel and L1-regularized SVM with linear kernel are available at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R}
and at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R},
respectively.

>>>>

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
diseases. However, the structure of the human microbiome is remarkably
variable between individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population individually. Recently,
machine learning models (ML) have grown in popularity among microbiome
researchers because of the large amount of data that can now be
generated and because the models are effective at accounting for the
interpersonal microbiome variation and the ecology of the disease.

ML models are useful for understanding the variation in the structure of
existing data and to apply that knowledge to make predictions about new
data. Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (1--16,
16--18). The task of diagnosing an individual with high confidence
relies on a model that is built with rigorous methods. However, there
are common methodological problems across many of these studies that
need to be addressed as the field progresses. These include a lack of
transparency in which methods are used and how these methods are
implemented; developing and evaluating models without a separate
held-out test data; large variation between the predictive performance
on different folds of cross-validation; and large variation between
cross-validation and testing performances. Nevertheless, the microbiome
field is making progress to avoid some of these pitfalls including
validating their models on independent datasets (7, 18, 19) and
introducing ways to better use ML tools (20--23). More work is needed to
further improve reproducibility and minimize over-optimism for model
performance.

@@

\textbf{Model selection and pipeline construction} We established a
common ML pipeline to train and validate each of the seven models that
is based on standard methods within the ML community (REFS by
Jenna?){[}Figure 1{]}.

@@

Second, we trained seven different models using the training data
{[}Table 1{]}. We selected models with different classification
algorithms and regularization methods. Regularization is a technique
that discourages overfitting by penalizing the model for learning the
training data too well. For regularized logistic regression and SVM with
linear kernel, we used L2-regularization to keep all potentially
important features. For comparison, we also trained an L1-regularized
SVM model with linear kernel. L1-regularization on microbiome data led
to a sparser solution (i.e., forced many coefficients to zero). To
explore the potential for non-linear relationships among features to
improve classification, we trained tree-based models including decision
tree, random forest, and XGBoost and we trained an SVM model with a
non-linear kernel.

Third, fitting of these models required selecting appropriate
hyperparameters. Hyperparameters are the rules that are learned from the
training set in a classification algorithm. For example, in the linear
models the regularization term (C) is a hyperparameter that indicates
the penalty for overfitting. Similar to regularization term C, all
hyperparameters are tuned to find the best model. We selected
hyperparameters by performing 100 five-fold cross-validation (CV)
repeats on the training set {[}Figure 1{]}. The five-fold CV was also
stratified to maintain the overall case and control distribution. We
chose the best hyperparameter values for each model based on its CV
predictive performance using the area under the receiver operating
characteristic curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC
ranges from 1.0, where the model perfectly distinguishes between cases
and controls, to 0.50, where the model's predictions are no different
from random chance. To select the best performing hyperparameter, we
performed a full grid search for hyperparameter settings when training
our models. Default hyperparameter settings in previously developed ML
packages in R, Python, and Matlab programming languages are inadequate
for effective application of classification algorithms and need to be
optimized for each new ML task. In the example of L1-regularized SVM
with linear kernel {[}Figure S1{]}, the model showed large variability
between different regularization coefficients (C) and was susceptible to
performing poorly if the wrong regularization coefficient was assigned
to the model by default.

Finally, we trained the full training dataset with the selected
hyperparameter values and applied the model to the held-out data to
evaluate the testing predictive performance of each model. The
data-split, hyperparameter selection, training and testing steps were
repeated 100 times to obtain a reliable and robust interpretation of
model performance {[}Figure 1{]}.

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had significantly higher test AUROC values than the
other models for detecting SRNs (Wilcoxon rank sum test, p \textless{}
0.01). The median AUROC of the random forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another and had median AUROC values of
0.680 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056), respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.650 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Interestingly, these results demonstrate that the most
complex model (XGBoost) did not have the best performance and that the
most interpretable models (L2-regularized logistic regression and SVM
with linear kernel) performed nearly as well as random forest.

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs was large, then that
would indicate that the models were overfit. The difference in median
AUROCs was 0.021 in L1-regularized SVM with linear kernel, followed by
SVM with radial basis function kernel and decision tree with a
difference of 0.007 and 0.006, respectively {[}Figure 2{]}. These
differences are relatively small and would not indicate a problem with
overfitting.

To evaluate the risk for over-optimism of each model, we calculated the
range of AUROC values for each model using 100 data-splits. The range
among the testing AUROC values within each model varied by 0.230 on
average across the seven models. If we had only done a single split,
then there is a risk that we could gotten lucky or unlucky with the
performance of the model. For instance, the lowest AUROC value of the
random forest model was 0.593 whereas the highest was 0.810. These
results showed that depending on the data-split, the testing AUROC
values showed great variability {[}Figure 2{]}. Therefore, it is
important to employ the 100 hierarchical data splits that were included
in our pipeline to minimize the risk of over-optimism.

@@

\textbf{Interpretation of each ML model.} Interpretability is the degree
to which humans can understand the reasons behind a model prediction
(31). Because we often use ML models not just to predict a health
outcome but also to identify the biomarkers for a disease, model
interpretation becomes crucial for microbiome studies. ML models
decrease in interpretability as they increase in complexity. In this
study we used two methods to help interpret our models.

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU was associated with classifying a subject as
being healthy or having an SRN. It was encouraging that many of the
highest ranked OTUs were shared across these three models, (e.g.~OTU 50,
426, 609, 822, 1239). The benefit of this approach was that the results
of the analysis were based on the trained model parameters and provided
information regarding the sign and magnitude of the impact of each OTU.
However, this approach is only possible with linear models.

Second, to analyze non-linear models we interpreted the feature
importance using permutation importance. Whereas the absolute feature
weights were determined from the trained models, here we measured
importance using the held-out test data. Permutation importance analysis
is a posthoc explanation of the model where we randomly permuted
non-correlated features individually and groups of perfectly correlated
features across the two groups in the held-out test data. We then
calculated how much the predictive performance of the model (i.e testing
AUROC values) decreased when each OTU or group of OTUs was randomly
permuted. We ranked the OTUs based on how much the median testing AUROC
decreased when it was permuted; the OTU with the largest decrease ranked
highest {[}Figure 4{]}. Among the twenty OTUs with the largest impact,
there was only one OTU (OTU 822) that was shared among all of the
models; however, we found three OTUs (OTU 58, 110, 367) that were
important in each of the tree-based models. Similarly, the random forest
and XGBoost models, shared four of the most important OTUs (OTU 2, 12,
361, 477). Permutation analysis results also revealed that with the
exception of the decision tree model, removal of any individual OTU had
a minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.680.
Effectively, the complexity of the communities was more fully
represented in the better performing models (22, 32). At least in this
case, it was not possible to distinguish between health and disease
using a single OTU. Although permutation analysis allowed us to gauge
the importance of an OTU, the analysis was post-hoc (i.e.~done using the
test data) and these results did not allow us to directly interrogate
the models to know whether an OTU was associated with classifying a
subject as being healthy or having an SRN.

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S4{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S4{]}, 17 of the 20 top OTUs (e.g.~OTU
609, 822, 1239) were deemed important by both interpretation methods.
Similarly, for the L2-regularized SVM and L2-regularized logistic
regression, 9 and 12 OTUs, respectively, were shared among the two
interpretation methods. Although permutation analysis does not not allow
us to determine the weight or the sign of the features, these results
indicate that both methods are consistent in selecting the most
important OTUs.

@@

There is a growing awareness that many human diseases and environmental
processes are not driven by a single organism but are the product of
multiple bacterial populations. Traditional statistical approaches are
useful for identifying those cases where a single organism is associated
with a process. In contrast, ML methods offer the ability to incorporate
the structure of the microbial communities as a whole to classify them
into different categories such as coming from a patient who is healthy
or has SRNs. If it is possible to classify communities reliably, then ML
methods also offer the ability to identify those microbial populations
within the communities that are responsible for the classification.
However, the application of ML in microbiome studies is still in its
infancy and the field needs to develop a better understanding of
different ML methods, their strengths and weaknesses, and how to
implement them.

To address these needs, we developed a framework to train rigorous,
transparent, and reproducible models. We benchmarked seven ML models and
showed that we can create models that are inherently interpretable and
easily trained without losing predictive performance. In terms of
predictive performance, the random forest model had the best testing
AUROC values compared to the other six models. However, the second-best
model was L2-regularized logistic regression with a median AUROC
difference of only 0.015 compared to random forest. While random forest
took 83.2 hours to train, L2-regularized logistic regression trained in
12 minutes. In terms of interpretability, random forest is a complex ML
model and could only be explained using post-hoc methods such as
permutation importance. On the other hand, L2-regularized logistic
regression was easier to interpret by ranking the OTUs based on their
feature weights and reviewing the signs of these weights. Comparing many
different models showed us that the most complex model was not
necessarily the best model for our ML task.

As we set out to select the best model, we established a pipeline that
can be generalized to any modeling method that uses 16S rRNA sequence
counts to predict a binary health outcome. We performed a random
data-split to create a training set (80\% of the data) and a held-out
test set (20\% of the data), which we used to evaluate predictive
performance. We repeated this data-split 100 times to measure the
possible variation in predictive performance. During the training, we
tuned the model hyperparameters with a repeated five-fold
cross-validation. Despite the high number of features microbiome
datasets typically have, the models we built with this pipeline were
generalizable as shown by the similar AUROC values from the
cross-validation and testing.

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures which create collinearity in the datasets. This can hinder
our ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (33). For example if one
OTU is highly correlated with another OTU, only one of these OTUs would
have a large feature weight thus hiding the importance of its correlated
OTU. To capture all important features, once we identify highly ranked
OTUs, we should review their relationships with other OTUs. These
relationships will help us generate new hypotheses about the ecology of
the disease and test them with follow-up experiments. When we used
permutation importance, we took collinearity into consideration by
grouping correlated OTUs to determine their impact as a group. We
grouped OTUs that had a perfect correlation with each other however, we
can reduce the correlation threshold to further investigate the
relationships among features. It is important to know the correlation
structures of the data to avoid misinterpreting the models. This is
likely to be a particular problem with shotgun metagenomic datasets
where collinearity will be more pronounced due to many genes being
correlated with one another because they come from the same chromosome.
To identify the true underlying microbial factors of a disease, it is
crucial to do correlation analyses and further experimentation for
biological validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with normal
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with normal colons and those with SRNs.
Furthermore, as number of categories to classify increase, more samples
are required for each category to train a model. We also did not use
regression-based analyses to predict a non-categorical outcome. We have
previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (34). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold great promise (11, 35, 36) but microbiome datasets
often suffer from having many features and small sample sizes, which
makes the deep learning models prone to overfitting. These methods are
even more complex than random forest and XGBoost and are considered
uninterpretable. There is great potential for applying ML approaches to
microbiome data and we can choose which ML approach is best for the
study.

@@

Our study highlights the need to make educated choices at every step of
developing a ML model with microbiome data. We created an aspirational
rubric that researchers can use to identify potential pitfalls when
using ML in microbiome studies and ways to avoid them {[}Table S1{]}. We
have highlighted the trade-offs between model complexity and
interpretability, the need for cross-validation to tune hyperparameters,
the utility of held-out test sets for evaluating predictive performance,
and the importance of considering correlation structures in datasets for
reliable interpretation. Furthermore, we underscored the importance of
proper experimental design and methods to help us achieve the level of
validity and accountability we want from models built for patient
health.

@@

\textbf{Data collection and study population.} The original stool
samples described in our analysis were obtained from patients recruited
by Great Lakes-New England Early Detection Research Network (Reference
from Mack?). Stool samples were provided by adults who were undergoing a
scheduled screening or surveillance colonoscopy. Participants were
recruited from: Toronto (ON, Canada), Boston (MA, USA), Houston (TX,
USA), and Ann Arbor (MI, USA). Patients' colonic health was visually
assessed by colonoscopy with bowel preparation and tissue histopathology
of all resected lesions. We assigned patients into two classes: those
with normal colons and those with screen relevant neoplasias (SRNs). The
normal class included patients with normal colons or non-advanced
adenomas whereas the SRN class included patients with advanced adenomas
or carcinomas (reference for SRN?). Patients with an adenoma greater
than 1 cm, more than three adenomas of any size, or an adenoma with
villous histology were classified as having advanced adenomas. There
were 172 patients with normal colonoscopies, 198 with adenomas, and 120
with carcinomas. Of the 198 adenomas, 109 were identified as advanced
adenomas. Together 261 patients were classified as normal and 229
patients were classified as having a SRN.

\textbf{16S rRNA gene sequencing data.} Stool samples provided by the
patients were used for 16S rRNA gene sequencing to measure bacterial
population abundances. The sequence data used in our analyses were
originally generated by Baxter et al. (available through NCBI Sequence
Read Archive {[}SRP062005{]}, 2015). The OTU abundance table was
generated by Sze et al (32), who processed the 16S rRNA sequences in
mothur (v1.39.3) using the default quality filtering methods,
identifying and removing chimeric sequences using VSEARCH, and assigning
to OTUs at 97\% similarity using the OptiClust algorithm (37--39);
(\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018/blob/master/data/process/baxter/baxter.0.03.subsample.shared}).
These OTU abundances were the features we used to predict colorectal
health of the patients. There were 6920 OTUs. OTU abundances were
subsampled to the size of the smallest sample and normalized across
samples such that the highest abundance of each OTU would be 1 and
lowest would be 0.

\textbf{Model training and evaluation.} Models were trained using the
caret package (v.6.0.81) in R (v.3.5.0). We modified the caret code to
calculate decision values instead of predicted probabilities for models
generated using L2-regularized SVM with linear kernel and L1-regularized
SVM with linear kernel. These changes were necessary to calculate AUROC
values for SVMs. The code for these changes on L2-regularized SVM with
linear kernel and L1-regularized SVM with linear kernel models are
available at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear3.R}
and at
\url{https://github.com/SchlossLab/Topcuoglu_ML_XXX_2019/blob/master/data/caret_models/svmLinear4.R},
respectively.

For hyperparameter selection, we started with a granular grid search.
Then we narrowed and fine-tuned the range of each hyperparameter. A full
grid search was needed to avoid large variation in prediction
performance. For L2-regularized logistic regression, L1 and
L2-regularized SVM with linear and radial basis function kernels, we
tuned the \textbf{cost} hyperparameter which determines the
regularization strength where smaller values specify stronger
regularization. For SVM with radial basis function kernel we also tuned
\textbf{sigma} hyperparameter which determines the reach of a single
training instance where for a high value of sigma, the SVM decision
boundary will be dependent on the points that are closest to the
decision boundary. For the decision tree model, we tuned the
\textbf{depth of the tree} where deeper the tree, the more splits it
has. For random forest, we tuned the \textbf{number of features} to
consider when looking for the best tree split. For XGBoost, we tuned for
\textbf{learning rate} and the \textbf{fraction of samples} to be used
for fitting the individual base learners. Recently developed tools such
as Hyperband help researchers with hyperparameter selection that can be
incorporated to microbiome studies (40) .

The computational burden during model training due to model complexity
was reduced by parallelizing segments of the ML pipeline. We
parallelized the training of each data-split. This allowed the 100
data-splits to be processed through the ML pipeline simultaneously at
the same time for each model. It is possible to further parallelize the
cross-validation step for each hyperparameter setting if limited by
computational resources.

\textbf{Permutation importance workflow.} We calculated a Spearman's
rank-order correlation matrix and defined correlated OTUs as having
perfect correlation (correlation coefficient = 1 and p \textless{}
0.01). Non-correlated OTUs were permuted individually whereas correlated
ones were grouped together and permuted at the same time.
